The ChEMBL database as Linked Open Data

### 2012-08-09 article.tex
Applied David's comments

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped  the ChEMBL dataset to the Resource Description Framework (RDF),
Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples, reflecting ``facts''. A fact can
be given on any topic, and RDF therefore talks about \textit{things} very generically~\cite{Miller:04:RP}. Things in RDF
are represented by Universal Resource Identifiers (URI) and when the more specialized
Universal Resource Locator (URL) is used, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not all, domains, making it a successful 
approach to data integration in the life sciences.

@@

A second resource in the linked open data network into which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF created a single RDF repository covering over 20 public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative searches~\cite{Chen2012a} and
data mining~\cite{Chen2012b}. The ChEMBL-RDF set has been uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL-RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link the two. For instance,
to investigate the relations between drug side effects and their targets (usually off-target effects), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart disease were linked to 87 unique protein targets ($IC_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and Sider in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.

>>>>

Using the right to download, modify, and redistribute the ChEMBL data, two independent teams
have previously mapped the ChEMBL dataset to the Resource Description Framework (RDF), resulting
in Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}. RDF is a framework where
knowledge is represented by small so-called triples, reflecting ``facts''. A fact can
be given on any topic, and RDF therefore talks about \textit{things} very generically~\cite{Miller:04:RP}. Things in RDF
are represented by Universal Resource Identifiers (URI) and when the more specialized
Universal Resource Locator (URL) is used, it creates \emph{linked data}~\cite{Samwald2011}. RDF is
developed as an open standard by the World Wide Web Consortium, and provides a knowledge
representation framework that applies to many, if not all, domains, making it a successful 
approach to data integration in the life sciences.

@@

A second resource in the linked open data network into which ChEMBL-RDF has been integrated is Chem2Bio2RDF.
Chem2Bio2RDF is a single RDF repository covering over twenty public data resources pertaining to drugs, chemical compounds, carcinogens,
protein targets, genes, diseases, side effects, pathways and their relations~\cite{Chen2010}. The entities and their relations were
further annotated by Chem2Bio2OWL ontology, making it a rich semantic resource for integrative searches~\cite{Chen2012a} and
data mining~\cite{Chen2012b}. The ChEMBL-RDF set has been uploaded into the Chem2Bio2RDF triple store, enabling
queries linking ChEMBL with other entities in Chem2Bio2RDF. Since both ChEMBL-RDF and Chem2Bio2RDF use InChI keys to present chemical
compounds and adopt Bio2RDF protein identifiers to present targets, queries can be easily constructed to link the two. For instance,
to investigate the relations between drug side effects and their targets (usually off-target effects), a query was created to link the
targets in ChEMBL to side effects (e.g., heart disease) in Chem2Bio2RDF via side effect related drugs and their bioassay activities.
In this case, 36 drugs causing heart disease were linked to 87 unique protein targets ($IC_50$ $<$ 10um). The top two most common targets
of these drugs are HERG and Serotonin transporter. HERG is a well-known off-target accounting for cardiovascular problems, while
the activity of Serotonin transporter may also be associated with significant cardiovascular adverse effects~\cite{Levy2006}. 
The targets can be also linked to side effects via protein-protein interactions by searching across ChEMBL, HPRD and Sider in
Chem2Bio2RDF. SPARQLs are available at \url{http://chem2bio2rdf.wikispaces.com/Integrate+with+ChEMBL} and can be executed in the
SPARQL endpoint~\cite{Chen2010}.
